Forecasting the age structure of the scientific workforce in Australia

### 2025-11-23 asf2.tex
Added diff

\begin{abstract}
Planning for a future workforce requires forecasts of age structure
changes to inform policy decisions, particularly related to universities
and immigration. We modify the traditional demographic growth-balance
equation to account for workforce-specific dynamics, replacing births
with graduate entry, modelling exits through death and retirement, and
including a remainder term that captures migration and career changes.
Functional data models are used to model age-specific components, while
ARIMA models are used for time series components. Simulation is employed
to generate forecast distributions, capturing uncertainty from all
components. The approach is illustrated using data on Australia's
scientific workforce, allowing us to forecast the age distribution of
various scientific disciplines for the next ten years. This analysis was
central to an Australian Academy of Science initiative examining the
capability of Australia's science system and identifying workforce gaps.
\end{abstract}

@@

@@

We propose to adapt this framework to workforce dynamics by redefining
the demographic components in two ways. First, we replace fertility with
workforce entry, which functions more like a migration process than a
birth process because graduates can enter the workforce at any age.
Second, we allow workers to leave the workforce through two processes:
retirement and death. Of course, people may also leave the workforce for
other reasons, such as a career change or family commitments, but since
we do not have data on these processes, we model them implicitly via a
remainder term.

@@

People can leave the workforce of a group through death, retirement,
emigration, or career change; they can enter the workforce through
graduation, immigration, or career change. Unfortunately, we typically
do not have data on many of these processes, so we will combine career
change, emigration and immigration into a remainder term, which we
denote as \(E_{i,x,t}\). Let \(D_{i,x,t}\) denote the number of deaths
of workers in group \(i\) of age \(x\) in year \(t\), \(R_{i,x,t}\)
denote the number of retirements from the same group of workers, and
\(G_{i,x,t}\) denote the number of new graduates of age \(x\) in year
\(t\) who take up work in group \(i\). The numbers in each case are for
people aged \(x\) at the \emph{start} of year \(t\). Then population
changes can be described using a model similar to the stochastic
population model of \citet{HB08}, excluding the birth process:
\begin{equation}\phantomsection\label{eq-dgb}{
P_{i,x+1,t+1} = P_{i,x,t} - D_{i,x,t} - R_{i,x,t} + G_{i,x,t} + E_{i,x,t},
}\end{equation} where

@@

@@

Functional data models have been widely used in demography and other
fields, and have been shown to work particularly well for age-specific
demographic processes \citep{HB08, LCcomparison}. They enable the
inherent smoothness over age to be captured, while modelling the
autocorrelation over time using relatively simple univariate time series
models applied to the principal component scores. In our application, we
use univariate ARIMA models for the \(q_{x,t}\) scores, and ARMA models
for the \(E_{i,x,t}\) scores. The assumption of stationarity for the
\(E_{i,x,t}\) scores is validated for the disciplines we consider.

For the time-varying component, \(G_{i,t}\), we use a global ARIMA model
\citep{HM21} to capture the dynamics over time and across disciplines.
The global model pools information across disciplines to improve
forecast accuracy, especially for disciplines with limited historical
data.

@@

@@

\begin{table}

@@

The remainder, \(E_{i,x,t}\), is also modelled using a functional data
model \citep{HU07}, with stationary ARMA models fitted to the principal
component scores. The scores were found to be stationary using the KPSS
test \citep{KPSS}. The forecasts for one year are shown in
Figure~\ref{fig-net-migration-disciplines}, with the mean forecast
represented by the solid line and 90\% prediction intervals indicated by
the shaded area.

>>>>

\begin{abstract}
Planning for a future workforce requires forecasts of age structure
changes to inform policy decisions, particularly related to universities
and immigration. We propose a new dynamic statistical model for
forecasting the age structure of a workforce. Our approach is inspired
by a stochastic model used in population forecasting, replacing births
with graduate entry, modelling exits through death and retirement, and
including a remainder term that captures migration and career changes.
Functional data models are used to model age-specific components, while
ARIMA models are used for time series components. Simulation is employed
to generate forecast distributions, capturing uncertainty from all
components. The approach is illustrated using data on Australia's
scientific workforce, allowing us to forecast the age distribution of
various scientific disciplines for the next ten years. This analysis was
central to an Australian Academy of Science initiative examining the
capability of Australia's science system and identifying workforce gaps.
\end{abstract}

@@

\urlstyle{tt}

@@

We propose a related approach for modelling workforce dynamics by
redefining the demographic components in two ways. First, we replace
fertility with workforce entry, which functions more like a migration
process than a birth process because graduates can enter the workforce
at any age. Second, we \emph{explicitly} model workers leaving the
workforce through two processes: retirement and death. Of course, people
may also leave the workforce for other reasons, such as a career change
or family commitments, but since we do not have data on these processes,
we model them \emph{implicitly} via a remainder term.

@@

People can leave the workforce of a group through death, retirement,
emigration, family responsibilities, or career change; they can enter
the workforce through graduation, immigration, changes in family
responsibilities, or career change. Unfortunately, we typically do not
have data on many of these processes, so we will combine changes due to
family responsibilities, career changes, emigration and immigration into
a remainder term, which we denote as \(E_{i,x,t}\). Let \(D_{i,x,t}\)
denote the number of deaths of workers in group \(i\) of age \(x\) in
year \(t\), \(R_{i,x,t}\) denote the number of retirements from the same
group of workers, and \(G_{i,x,t}\) denote the number of new graduates
of age \(x\) in year \(t\) who take up work in group \(i\). The numbers
in each case are for people aged \(x\) at the \emph{start} of year
\(t\). Then population changes can be described using the following
model: \begin{equation}\phantomsection\label{eq-dgb}{
P_{i,x+1,t+1} = P_{i,x,t} - D_{i,x,t} - R_{i,x,t} + G_{i,x,t} + E_{i,x,t},
}\end{equation} where

@@

While our model was inspired by the stochastic population model of
\citet{HB08}, that model has different inputs (births and immigration)
and fewer outputs (deaths and emigration). Labour market forecasting is
more complicated with no birth process, several more inputs (graduates,
immigration, career changes, career renewal), and several more outputs
(deaths, retirements, emigration, career disruption and career changes).

@@

Because the \(\beta_{k,t}\) values are principal component scores, they
are uncorrelated by construction. While it is possible for there to be
some cross-correlations between the series at lags other than zero,
these are usually not large enough for a multivariate model to give more
accurate forecasts \citep[see,][]{HU07} except in contrived simulated
examples. On the other hand, \citet{Aue2015} did use multivariate models
to capture these cross-correlations, although they did not compare them
on real data.

Functional data models have been widely used in demography and other
fields, and have been shown to work particularly well for age-specific
demographic processes \citep{HB08, LCcomparison}. They enable the
inherent smoothness over age to be captured, while modelling the
autocorrelation over time using relatively simple univariate time series
models applied to the principal component scores. In our application, we
use univariate ARIMA models for the \(q_{x,t}\) scores, and ARMA models
for the \(E_{i,x,t}\) scores, each estimated using maximum likelihood
estimation. The assumption of stationarity for the \(E_{i,x,t}\) scores
is validated for the disciplines we consider, but is not a requirement
in general.

For the time-varying component, \(G_{i,t}\), we use a global ARIMA model
\citep{HM21} to capture the dynamics over time and across disciplines.
This is estimated using least squares estimation. The global model pools
information across disciplines to improve forecast accuracy, especially
for disciplines with limited historical data.

@@

Fortunately, there is no reason to think scientists of different
disciplines would have different mortality experiences. A century ago,
the dangers of radiation did increase mortality rates amongst chemists
and physicists compared to other sciences, but modern science is
conducted in extremely safe environments, so it seems reasonable to
assume that all science disciplines share similar mortality profiles.

There has been a small increase in average retirement age over the last
ten years due to an increase in the age at which the old age pension can
be accessed \citep{pension}, and a steady increase in the preservation
age at which superannuation can be accessed
\citep{kingston2019superannuation}. However, there is no existing policy
proposal to change either of these in the future, so it is reasonable to
take the retirement age distribution in recent years as valid for the
foreseeable future. Further, we know of no evidence that the
socio-economic status of scientists varies with discipline, so there is
no reason to think retirement intentions would change with discipline
either.

We assume the age distribution of graduates is a product of
age-dependent and time-dependent variables, \(g_x\) and \(G_{i,t}\).
Primarily, this is a pragmatic choice because we do not have more
detailed data available. We can get age distributions of graduates
across all disciplines in Australia, but not for each discipline; and we
can get the numbers of graduates by discipline and year in Australia,
but with no age breakdown. The most likely consequence of this
simplifying assumption is that the variability in graduate numbers by
age and time could be underestimated. It is conceivable that older
graduates are drawn to different disciplines than younger graduates, or
that fashionable disciplines change over time, resulting in different
age distributions of the graduates over time. But without specific data
related to this issue, we can only speculate.

It is also worth pointing out that the remainder term \(E_{i,x,t}\) will
absorb any inaccuracies that result from simplifying model assumptions
in the other components, and we forecast the remainder allowing for
changes over time, age and discipline. In fact, we could ignore all the
model components and just forecast \(P_{i,x,t}\) directly using a
functional time series model, but that would fail to separate out the
competing dynamics at play, and lead to much wider prediction intervals.
By trying to model the individual components where we have available
data, even if imperfectly, we capture more of the inherent uncertainty
and obtain narrower prediction intervals.

@@

\begin{table}[!hb]

@@

The remainder, \(E_{i,x,t}\), is also modelled using a functional data
model \citep{HU07}, with ARIMA models fitted to the principal component
scores. In this case, all scores were found to be stationary using the
KPSS test \citep{KPSS}, so ARMA models are used. The forecasts for one
year are shown in Figure~\ref{fig-net-migration-disciplines}, with the
mean forecast represented by the solid line and 90\% prediction
intervals indicated by the shaded area.
